---
title: "p8105_HW2_yp2846"
output: github_document
---
# Problem 1

```{r}
library(tidyverse)
transit_df = 
  read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           na =c("NA","."," "))|>
  janitor::clean_names() |>
  select(
    line, 
    station_name, 
    station_latitude, 
    station_longitude, 
    route1:route11, 
    entry, 
    vending, 
    entrance_type, 
    ada)|>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
  
dim(transit_df)
head(transit_df,10)
```
The cleaned subway entrance dataset contains 1868 rows(observations) and 19 columns(variables). 

The variables include subway line and station name, station latitude and longitude, routes served, whether entry is allowed, whether vending is available, entrance type, and ADA compliance. 

I cleaned the variable names using janitor::clean_names(), retained only the variables relevant to the analysis and converted the entry variable from "YES" and "NO" character values to a logical variable with TRUE and FALSE. 

The resulting dataset is not completely tidy because the routes served are stored across multiple columns, from route1 through route11 rather than in a single route variable.

```{r}
transit_df |> 
  distinct(line, station_name) |>
  nrow()
```

There are `r nrow(distinct(transit_df,line,station_name))` distinct stations.

```{r}
transit_df |>
  filter(ada == TRUE)|>
  distinct(line, station_name) |>
  nrow()
```

There are 84 ADA compliant.

```{r}
transit_df |>
  filter(
    vending == "NO") |>
  summarize(
    transit_proportion = mean(entry)
  )
```

The proportion of station entrances/ exits without vending allow entrance is about 0.3770.

```{r}
transit_long =
  transit_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE,
    values_transform = as.character
  )

transit_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```

There are 60 distinct stations serve the A train.

```{r}
transit_long |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()
```

Of the stations that serve the A train, there are 17 are ADA compliant.


