---
title: "p8105_HW2_yp2846"
output: github_document
---
# Problem 1

```{r}
library(tidyverse)
transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           na =c("NA","."," "))|>
  janitor::clean_names() |>
  select(
    line, 
    station_name, 
    station_latitude, 
    station_longitude, 
    route1:route11, 
    entry, 
    vending, 
    entrance_type, 
    ada)|>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
  
dim(transit_df)
head(transit_df,10)
```
The cleaned subway entrance dataset contains 1868 rows(observations) and 19 columns(variables). 

The variables include subway line and station name, station latitude and longitude, routes served, whether entry is allowed, whether vending is available, entrance type, and ADA compliance. 

I cleaned the variable names using janitor::clean_names(), retained only the variables relevant to the analysis and converted the entry variable from "YES" and "NO" character values to a logical variable with TRUE and FALSE. 

The resulting dataset is not completely tidy because the routes served are stored across multiple columns, from route1 through route11 rather than in a single route variable.

```{r}
transit_df |> 
  distinct(line, station_name) |>
  nrow()
```

There are `r nrow(distinct(transit_df,line,station_name))` distinct stations.

```{r}
transit_df |>
  filter(ada == TRUE)|>
  distinct(line, station_name) |>
  nrow()
```

There are 84 ADA compliant.

```{r}
transit_df |>
  filter(
    vending == "NO") |>
  summarize(
    transit_proportion = mean(entry)
  )
```

The proportion of station entrances/ exits without vending allow entrance is about 0.3770.

```{r}
transit_long =
  transit_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE,
    values_transform = as.character
  )

transit_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```

There are 60 distinct stations serve the A train.

```{r}
transit_long |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()
```

Of the stations that serve the A train, there are 17 are ADA compliant.

# Problem 2

```{r}
library(readxl)
```

```{r}
# Read and clean the Mr. Trash Wheel sheet
trash_Mr_df =
  read_excel("./data/202509 Trash Wheel Collection Data.xlsx",
             sheet = "Mr. Trash Wheel",
             range = "A2:N709")|>
  janitor::clean_names()|>
  mutate(
    as.integer(round(sports_balls))
  )|>
  mutate(
    year = as.integer(year),
    trash_wheel = "Mr."
  )
# head(trash_df, 5)
# tail(trash_df, 5)
dim(trash_Mr_df)
```

```{r}
# Read and clean the Professor Trash Wheel sheet
trash_professor_df =
  read_excel("./data/202509 Trash Wheel Collection Data.xlsx",
             sheet = "Professor Trash Wheel",
             range = "A2:M134")|>
  janitor::clean_names()|>
  mutate(
    year = as.integer(year),
    trash_wheel = "Professor"
  )
# head(trash_professor_df, 5)
# tail(trash_professor_df, 5)
dim(trash_professor_df)
```

```{r}
# Read and clean the Gwynns Falls Trash Wheel sheet
trash_Gwynns_df =
  read_excel("./data/202509 Trash Wheel Collection Data.xlsx",
             sheet = "Gwynns Falls Trash Wheel",
             range = "A2:L351")|>
  janitor::clean_names()|>
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynns"
  )
# head(trash_Gwynns_df, 5)
# tail(trash_Gwynns_df, 5)
dim(trash_Gwynns_df)
```

```{r}
trash_combine = 
  bind_rows(
    trash_Mr_df,
    trash_professor_df,
    trash_Gwynns_df
  )
# head(trash_combine,5)
# tail(trash_combine,5)
# dim(trash_combine)
```

There are `r nrow(trash_combine)` rows and `r ncol(trash_combine)` columns in resulting dataset of sheet "Mr. Trash Wheel","Professor Trash Wheel" and "Gwynns Falls Trash Wheel".

Key variables are `r colnames(trash_combine)`.

The total weight of trash collected by Professor Trash Wheel is `r sum(trash_professor_df$weight_tons, na.rm=TRUE)`.

The total number of igarette butts collected by Gwynnda in June of 2022 is `r trash_Gwynns_df |> filter(month == "June" & year == 2022) |> summarize( sum(cigarette_butts, na.rm=TRUE))|>pull()|>format(scientific = FALSE)`. 

# Problem 3
```{r}
baseline_df =
  read_csv("./data/data_mci/MCI_baseline.csv",
                        skip = 1,
                        na = c("NA"," ","."))|>
  janitor::clean_names()|>
  mutate(
    sex = recode(
      sex,
      '0' = "Female",
      '1' = "Male"
    ),
    apoe4 = recode(
      apoe4,
      '0' = "APOE4 non-carrier",
      '1' = "APOE4 carrier"
    )
  )
head(baseline_df,5)
tail(baseline_df,5)



baseline_df =
  read_csv("./data/data_mci/MCI_baseline.csv",
                        skip = 1,
                        na = c("NA"," ","."))|>
  janitor::clean_names()
head(baseline_df)
```

```{r}
baseline_df |>
  filter(is.na(age_at_onset)| age_at_onset > current_age)
```

```{r}
baseline_df |>
  summarize(
    n_participants = n(),
    n_mci = sum(!is.na(age_at_onset)),
    mean_age = mean(age_at_onset),
    proportion_female =sum(sex=="Female" & apoe4 == "APOE4 carrier")/sum(sex=="Female") 
  )
```



The variables include `r colnames(baseline_df)`. 

I cleaned the variable names using janitor::clean_names(), retained only the variables relevant to the analysis. 
I converted the sex variable from "0" and "1" double values to a character variable with "Female" and "Male". 
I also converted apoe4 variable from "0" and "1" double values to character variable with "APOE4 non-carrier" and "APOE4 carrier".




```{r}
amyloid_df =
  read_csv("./data/data_mci/mci_baseline.csv",
                        skip = 1,
                        na = c("NA"," ","."))|>
  janitor::clean_names()|>
  mutate(
    sex = as.logical(sex),
    apoe4 = as.logical(apoe4)
  )
head(amyloid_df,5)
tail(amyloid_df,5)
```


